---
title: 'R Course: Lesson 4'
author: "Page Piccinini"
date: "March 17, 2016"
output: html_document
---
```{r, echo=FALSE}
load("rcourse_lesson4_environment.RData")

library(knitr)
```

# Introduction

Today I looked at election data from eight United States presidential elections (1964, 1972, 1980, 1984, 1992, 1996, 2004, 2012). Specifically, I looked at elections where an incumbent was running for president. I wanted to see if the percentage of the population that voted for the incumbent varied by the political party of the incumbent (Democrat, Republic) and whether the state was party of the Union or the Confederacy during the Civil War.

# Data

Since I'm using an ANOVA today, I need to make sure my data set was balanced. So, instead of taking all states that were offically states during the Civil War I made sure it was the same number in each group (Union, Confederacy). There were only 11 Confederacy states, so to get a matched sample of Union states I used for the first 11 states that were admitted to the Union. For example, California was not included because it join the Union later.

# Results

First I plotted histograms of our data separated by our two varaibles to be sure they had a normal distribution. I did this over the entire data set and separating the data by my categorical varaibles. The distributions do not look completely normal (particularly for the summarised data), but enough such that I am going to more forward with they analysis without transforming my data.

```{r, echo=FALSE, fig.align='center'}
incumbent_histogram_full.plot
incumbent_histogram_sum.plot
```

I've plotted my data both as a boxplot and a barplot, but I think the boxplot is better. Both suggest that states that stayed in the Union vote more for incumbent Democrats than incumbent Republicans, conversely, states that joined the Confederacy show the opposite effect. This suggests there may be an interaction in the data.

```{r, echo=F, fig.align = 'center'}
incumbent_boxplot.plot
incumbent_barplot.plot
```

To test this effect I ran an analysis of variance (ANOVA). The dependent variable was percentage of the population voting for the incumbent, the indepedent variables were incumbent party (Democrat, Republican) and the state's country during the Civil War (Union, Confederacy). Incumbent party was included as a within-state variable. The model did not find either main effect to be significant, although the effect of incumbent party was trending [F(1, 20) = 3.59, p = 0.07]. There was a significnat intearction of incumbent party and civil war.

```{r}
incumbent.aov_sum
```

To better understand the interaction of incumbent party and civil war, I ran t-tests looking at my two main effects within subsets of the data. Looking first within civil war country, I ran a paired t-test to see if both groups showed a difference of incumbent party. I find that for union states there is a significant effect at p < 0.05 (p = `r incumbent_union.ttest$p.value`), but this will not hold with p-value correction. However, for states from the confederacy the effect remains very strong (p = `r incumbent_confederacy.ttest$p.value`), showing that confederacy states have a strong preference for Republican incumbents.

```{r}
incumbent_union.ttest
incumbent_confederacy.ttest
```

I can also look at the results within party, to see if there is a difference depending on the status of a state during the Civil War. To do this I ran an unpaired t-test within each party. For Democrats there is clearly an effect (p = `r incumbent_democrat.ttest$p.value`), where Democratic incumbents get a much higher percentage of votes from Union states than Confederacy state. Republicans show a strong effect in the reverse direction (p = `r incumbent_republican.ttest$p.value`), with getting a higher percentage of votes from Confederacy states.

```{r}
incumbent_democrat.ttest
incumbent_republican.ttest
```

# Conclusion

A states' status during the Civil War has a large effect on how it votes. However, the effect is stronger for states from the Confederacy, who have a much stronger Republican bias than Union states have a Democrat bias.